Ecosystem size and complexity dictate riverine biodiversity

### 2021-03-14 manuscript_sciad.Rmd
edits: remove methods

Here, we show that ecosystem size and complexity play comparable roles in controlling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

@@

First, we developed a theoretical framework synthesizing the influences of ecosystem size and complexity on biodiversity patterns in branching ecosystems. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit [@yeakelSynchronisationStabilityRiver2014; @teruiMetapopulationStabilityBranching2018] (**Figure 1A and B**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Materials and Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity model [@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Materials and Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementary Text**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation (**Figure 1A**). This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries but are highly correlated within a tributary [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to limited influences of branching on habitat heterogeneity (**Figure 1B**). This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities [@poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel $\theta$), dispersal probability $p_{d}$, and the maximum value of interspecific competition strength $\alpha_{max}$ (see **Materials and Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Text**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-S11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity [@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-S18**). Under this scenario, branching has a minor influence on the ecosystem's habitat heterogeneity because patch-level environmental variation is equivalent to or greater than environmental differences between tributaries (**Figure 1B**). Therefore, this landscape scenario eliminates the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams show distinct and spatially-correlated environmental conditions, including water temperature, water chemistry, and flow/sediment regimes [@petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015]. Instead, it may be most relevant to severely altered landscapes where localized human disturbance disrupts the environmental distinctiveness of branches through, for example, flow regulations by dams [@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

@@

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions. The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supplementary Text**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016] - the sole theoretical assumption required for the dual control of $\gamma$ diversity to emerge. Further, the consistent patterns across geographic regions with distinct fish fauna suggest that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

@@

>>>>

Here, we show that ecosystem size and complexity play comparable roles in controlling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

@@

First, we developed a theoretical framework synthesizing the influences of ecosystem size and complexity on biodiversity patterns in branching ecosystems. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit [@yeakelSynchronisationStabilityRiver2014; @teruiMetapopulationStabilityBranching2018] (**Figure 1A and B**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Materials and Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity [@mcguireNetworkAnalysisReveals2014]. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity model [@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Materials and Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementary Text**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation (**Figure 1A**). This landscape scenario reproduces natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries but are highly correlated within a tributary[@petersonModellingDendriticEcological2013; @mcguireNetworkAnalysisReveals2014]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to limited influences of branching on habitat heterogeneity (**Figure 1B**). This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities [@poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel $\theta$), dispersal probability $p_{d}$, and the maximum value of interspecific competition strength $\alpha_{max}$ (see **Materials and Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Text**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-S11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity [@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-S18**). Under this scenario, branching has a minor influence on the ecosystem's habitat heterogeneity because patch-level environmental variation is equivalent to or greater than environmental differences between tributaries (**Figure 1B**). Therefore, this landscape scenario eliminates the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams show distinct and spatially-correlated environmental conditions, including water temperature, water chemistry, and flow/sediment regimes [@petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @mcguireNetworkAnalysisReveals2014]. Instead, it may be most relevant to severely altered landscapes where localized human disturbance disrupts the environmental distinctiveness of branches through, for example, flow regulations by dams [@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

@@

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions. The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supplementary Text**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual control of $\gamma$ diversity to emerge. Further, the consistent patterns across geographic regions with distinct fish fauna suggest that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

@@

# Materials and Methods

## Theory

### Network generation

We depicted branching ecosystems as a bifurcating network of connected habitat patches [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014] (**Figure 1**). Habitat patches, or local communities, can be either non-branching or branching river sections with a unit length $l$, which defines the scale of local species interactions. Two parameters determine the geometric properties of simulated branching networks: the number of habitat patches $N_p$ (ecosystem size) and branching probability $P_b$ (ecosystem complexity). Each habitat patch is assigned to be a branching patch (including upstream terminals) with probability $P_b$ or a non-branching patch with probability $1-P_b$. In this framework, an individual branch is a consecutive series of non-branching patches terminated at a branching patch; therefore, the number of habitat patches in branch $q$, $n_{p,q}$, is a realization of a random variable drawn from a geometric distribution $n_{p,q} \sim Ge(P_{b})$ [@teruiMetapopulationStabilityBranching2018] but conditional on $\sum_{q=1}^{N_b}n_{p,q} = N_p$ ($N_b$ is the number of branches in a network). The number of branches $N_b$ is identical to the number of branching habitat patches in a network, or the number of 'success' out of $N_p$ trials. This leads to an assumption that $N_b$ is a realization of a random variable drawn from a binomial distribution as $N_b \sim Binomial(N_p, P_b)$. This framework has two merits. First, it reflects observed patterns of branch length distribution, which is known to follow an exponential distribution (a continuous version of a geometric distribution) [@rodriguez-iturbeFractalRiverBasins2001]. Second, it preserves the fractal nature of branching patterns [@peckhamReformulationHortonLaws1999]. An alternative approach is the use of optimal channel networks [@carraroGenerationApplicationRiver2020a]. However, we employed our framework because branching probability $P_b$ is quantifiable in natural river networks, enabling us to compare theoretical predictions with empirical observations.

In our network generation, we first drew $N_b$ from the binomial distribution but selected an odd number to constitute a bifurcating branching network. We then determined $n_{p,q}$ as random draws from the geometric distribution with the constraint of $\sum_{q=1}^{N_b}n_{p,q} = N_p$. The generated branches were organized randomly into a bifurcating branching network.

Two sources of variation characterize the long-term average of the abiotic environment at each habitat patch: environmental variation at headwaters ($\sigma_h$) and local environmental noise ($\sigma_l$). We drew random values from a normal distribution with a mean of zero and a standard deviation of $\sigma_{h}$ and assigned them to headwater patches (i.e., the most upstream patches). These environmental values at headwaters propagate downstream through a spatial autoregressive process defined as:

```{=tex}
\begin{equation}
  \mu_{z,k} = \rho \mu_{z,k+1} + \epsilon_{k}
\end{equation}
```
where $\mu_{z,k}$ is the mean environmental value at longitudinal position $k$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet), $\rho$ is the strength of spatial autocorrelation and $\epsilon_k$ is the local environmental noise that follows a normal distribution with a mean of zero and a standard deviation of $\sigma_l$. The parameter $\rho$ can take values of 0-1 with larger values indicating greater spatial autocorrelation. In this study, we set $\rho = 1$ to mimic strong spatial autocorrelation in riverine environments. At confluences, we took a weighted mean of environmental values given the relative size of upstream contributing area of joining tributaries:

```{=tex}
\begin{equation}
\begin{aligned}
  \mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) \\
  \omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L}
\end{aligned}
\end{equation}
```
where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area; $A_k = 1$ at the most upstream patch). The parameter $\omega$ takes values of 0-1 and represents the relative influence of the right tributary given the relative size of joining tributaries. With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004].

### Metacommunity model

We simulated metacommunity dynamics in branching river networks following the metacommunity framework proposed by Thompson et al [@thompsonProcessbasedMetacommunityFramework2020]. In this framework, abiotic environmental conditions (density-independent), intra- and interspecific competition (density-dependent), and dispersal regulate a species' realized population growth. Below, we describe the details of our metacommunity model.

The realized number of individuals $N_{ix}(t + 1)$ (species $i$ at patch $x$ and time $t+1$) is given as:

```{=tex}
\begin{equation}
  N_{ix}(t + 1) \sim Poisson(n_{ix}(t) + I_{ix}(t) - E_{ix}(t))
\end{equation}
```
where $n_{ix}(t)$ is the expected number of individuals given the local community dynamics at time $t$, $I_{ix}(t)$ the expected number of immigrants to patch $x$, and $E_{ix}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes [@fronhoferClassicalMetapopulationDynamics2017; @thompsonProcessbasedMetacommunityFramework2020]. Local community dynamics are simulated based on the Beverton-Holt equation [@bevertonDynamicsExploitedFish1957]:

```{=tex}
\begin{equation}
  n_{ix}(t) = \frac{N_{ix}(t)r_{ix}(t)}{1 + \frac{r_{0,i}-1}{K_{x}}\sum_{j=1}^S{\alpha_{ij}N_{jx}(t)}}
\end{equation}
```
where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. $K_x$ was expressed as a function of the number of upstream habitat patches $A_x$ ($K_x = 100A_x$) to represent increasing habitat size with increasing upstream watershed area [@teruiThreeEcologicalFactors2016; @rahelFishAssemblagesHabitat1991]. The parameter $\alpha_{ij}$ represents the strength of interspecific competition relative to that of intraspecific competition: interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ \> 1 (intraspecific competition coefficient was set constant as $\alpha_{ii} = 1$). $\alpha_{ij}$ was drawn randomly from a uniform distribution as $\alpha_{ij} \sim Unif(0, \alpha_{max})$. The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

```{=tex}
\begin{equation}
  r_{ix}(t) = \delta~r_{0,i}~exp[{-\frac{\{\mu_i-z_x(t)\}^2}{2\sigma^2_{niche,i}}}]
\end{equation}
```
where $\mu_i$ is the optimal environmental value for species $i$, $z_x(t)$ the environmental value at patch $x$ and time $t$, and $\sigma_{niche,i}$ the niche width of species $i$. We introduced the cost of a wide niche by multiplying $\delta$ [@chaianunpornEvolutionaryResponsesClimate2015]:

```{=tex}
\begin{equation}
  \delta = exp(-\frac{\sigma_{niche,i}^2}{2\nu^2})
\end{equation}
```
Smaller values of $\nu$ imply greater costs of having a wider niche (i.e., decreased maximum reproductive rate) and reduce the realized reproductive rate $r_{ix}(t)$ near the niche optimum.

The environmental value at patch $x$ and time $t$, $z_x(t)$, is assumed to follow a multivariate normal distribution as $z_{x}(t) \sim MVN(\boldsymbol{\mu_z}, \boldsymbol{\Omega_z})$, where $\boldsymbol{\mu_z}$ is the vector of mean environmental conditions of habitat patches and $\boldsymbol{\Omega_z}$ is the variance-covariance matrix. The diagonal elements of the variance-covariance matrix $\sigma^2_z$ represent the degree of temporal environmental variation. The off-diagonal elements $\Omega_{xy}$ represent the spatial autocorrelation in temporal environmental variation as:

```{=tex}
\begin{equation}
  \Omega_{xy} = \sigma_z^2~exp(-\phi d_{xy})
\end{equation}
```
$\Omega_{xy}$ denotes the temporal covariance of environmental conditions between patch $x$ and $y$, which is assumed to decay exponentially with increasing distance between the patches $d_{xy}$ (the shortest network path measured as the number of habitat patches). The parameter $\phi$ determines the degree of the distance decay of environmental correlates.

The expected number of emigrants at time $t$ is the product of dispersal probability $p_d$ and $n_{ix}(t)$: $E_{ix}(t) = p_{d}n_{ix}(t)$. The immigration probability at patch $x$ for species $i$, $\xi_{ix}(t)$, is calculated using the following equation that accounts for separation distance among habitat patches and dispersal capability of a species:

```{=tex}
\begin{equation}
  \xi_{ix}(t) = \frac{\sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}{\sum_{x} \sum_{y,y \neq x}E_{iy}(t)~exp(-\theta d_{xy})}
\end{equation}
```
The parameter $\theta$ regulates the dispersal distance of a species that follows an exponential distribution ($\theta^{-1}$ corresponds to the expected dispersal distance). The expected number of immigrants is calculated as $I_{ix}(t) = \xi_{ix}(t)\sum_x^{N_p} E_{ix}$.

### Simulation

We used `r n_scenario` sets of parameter combinations (i.e., simulation scenarios) based on the results of extensive sensitivity analysis (**Supplementary Text**). The parameter combinations comprises four landscape and eight ecological scenarios. Landscape scenarios distinguish spatial patterns of habitat heterogeneity, while ecological scenarios change ecological traits of simulated species. For landscape scenarios, we used four combinations of environmental variation at headwaters ($\sigma_h = 0.01, 1$) and the degree of local environmental noise ($\sigma_l = 0.01, 1$). The inequality of $\sigma_h$ and $\sigma_l$ controls the relationship between branching complexity and habitat heterogeneity (**Figures 1A and 1B**). For ecological scenarios, we varied three ecological parameters that are relevant for dispersal and interspecific competition: dispersal distance ($\theta =$ `r unique(dat_sim$theta)`), dispersal probability ($p_d =$ `r unique(dat_sim$p_dispersal)`), and the maximum value of competition coefficients ($\alpha_{max} =$ `r unique(dat_sim$max_alpha)`). This setup results in eight ecological scenarios with different levels of dispersal and competition.

We allowed interspecific variation in niche optimum and width, which were drawn from uniform distributions as $\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$. We used fixed values for the following parameters: maximum reproductive rate ($r_{0,i} =$ 4), niche cost ($\nu =$ `r unique(dat_sim$niche_cost)`), the degree of temporal environmental variation ($\sigma_z =$ `r unique(dat_sim$sd_env)`), and the extent of spatial autocorrelation in temporal environmental dynamics ($\phi =$ `r unique(dat_sim$phi)`).

Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 200 time steps for initialization and 200 time steps for burn-in) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`). During the first 200 time steps (an initialization period), we initialized the simulation by seeding each habitat patch with populations of each species drawn from a Poisson distribution with a mean of 0.5. We repeated the seeding every 10 time steps during the initialization to allow species to establish their populations if they can grow from low abundances. After the initialization period, we ran 200 time steps as a burn-in period with no seeding to minimize influences of initial conditions. We saved the last 1000 time steps to estimate the temporal averages of $\alpha$, $\beta$, and $\gamma$ diversity. We removed simulation replicates in which no species established populations over the initial 400 time steps (`r sum(dat_sim$alpha_div == 0)` replicates out of `r nrow(dat_sim)` replicates). Ecosystem size and complexity were drawn randomly from uniform distributions as: $N_p \sim Unif(10, 150)$ and $P_b \sim Unif(0.01, 0.99)$ ($N_p$ was rounded to the nearest integer before running simulations). Values of simulation parameters were summarized in **Table S2**. We performed simulations using R version 4.0.2 [@rcoreteamLanguageEnvironmentStatistical2020].

## Empirical data

### Fish community data

We compiled fish data at two geographic regions: the Hokkaido island, Japan, and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish [@fukushimaModellingEffectsDams2007; @fukushimaHokkaidoFreshwaterFish2011], monitoring data at protected watersheds [@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors [@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire island. For the Midwest, we assembled fish community data collected by the Iowa Department of Natural Resources [@iowadepartmentofnaturalresourcesBiologicalAssessmentIowa2004], Illinois Environmental Protection Agency and Illinois Department of Natural Resources [@illinoisenvironmentalprotectionagencyIllinoisWaterMonitoring2014], Minnesota Pollution Control Agency [@minnesotapollutioncontrolagencyDevelopmentFishbasedIndex2014], and Wisconsin Department of Natural Resources [@wisconsindepartmentofnaturalresourcesGuidelinesAssessingFish2018]. We detailed procedures for initial data selection in **Supplementary Text**.

We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). Metacommunities with $\ge$ 10 sampling sites were used in our analysis because a small sample size inflates statistical uncertainties in estimating asymptotic species richness (**Supplementary Text**).

We estimated $\alpha$, $\beta$, and $\gamma$ diversity at each watershed. $\gamma$ diversity is the total species richness at each watershed. Since it is impossible to sample all species present in each watershed, we used asymptotic species richness as a proxy for $\gamma$ diversity. Asymptotic species richness can be interpreted as the estimated true species richness in a given area when estimated using spatial replicates of local community data within a metacommunity [@chaoRarefactionExtrapolationHill2014a]. We estimated asymptotic species richness using sampling sites located in the same watershed (i.e., spatial replicates within a metacommunity) via the R package 'iNEXT' [@hsiehINEXTPackageRarefaction2016]. We assessed the completeness of the observed species richness based on sample coverage, which is the total proportional abundances of observed species to the whole metacommunity [@chaoRarefactionExtrapolationHill2014a]. We excluded watersheds whose sample coverage was below 0.9 because the estimates of asymptotic species richness may contain non-negligible statistical uncertainty. $\alpha$ diversity was estimated by taking an average of local species richness at each site. $\beta$ diversity was estimated by dividing the estimated $\gamma$ diversity by $\alpha$ diversity [@whittakerVegetationSiskiyouMountains1960].

After the data selection procedure, we obtained diversity estimates from `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (Hokkaido: `r nrow(wsd_subset[[1]])`, Midwest: `r nrow(wsd_subset[[2]])`), which encompassed a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites (Hokkaido: `r nrow(point_subset[[1]])`, Midwest: `r nrow(point_subset[[2]])`). Observations ranged from `r min(dat_hkd_site$Year, na.rm = T)` to `r max(dat_hkd_site$Year, na.rm = T)` in Hokkaido and from `r min(dat_mw_site$Year, na.rm = T)` to `r max(dat_mw_site$Year, na.rm = T)` in the Midwest. Fish species observed in the final data set were listed in **Tables S5 and S6**.

### Environmental data

We collected data for watershed characteristics (watershed area [km^2^], branching probability, and mean elevation [m]), climatic variables (mean annual air temperature [degree Celsius] and cumulative precipitation [mm]), and land use (the fractions of forest, urban, and agriculture, and dam density per watershed area [km^-2^]). We used Hydrologically Adjusted Elevations of MERIT Hydro [@yamazakiMERITHydroHighResolution2019] to delineate watershed polygons and river lines (defined as $\ge$ 1 km^2^ in the watershed area), which were separated by either the ocean, large lakes/reservoirs ($\ge$ 10 km^2^ in the areal area), or large rivers ($\ge$ 5000 km^2^ in the watershed area). The ocean and large lakes/reservoirs were extracted from Global 1-second Water Body Map ver 1.0 [@yamazakiDevelopmentGlobal90m2015]. Watershed area was estimated based on the delineated watershed polygons. We estimated branching probability by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We obtained the mean elevations for each watershed polygon using Hydrologically Adjusted Elevations of MERIT Hydro [@yamazakiMERITHydroHighResolution2019]. We estimated the spatial averages of mean annual air temperature and cumulative precipitation for each watershed (\~1 km resolution), which were extracted from WorldClim 1.4 [@hijmansVeryHighResolution2005]. We calculated the fractions of forest, urban, and agriculture for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution) [@marcelbuchhornCopernicusGlobalLand2020]. We calculated dam density per watershed area (km^-2^) based on the Global Reservoir and Dam Database, Version 1 (GRanDv1) [@lehnerGlobalReservoirDam2011]. We used ArcMap 10.7, SAGA 6.3.0, and the following R packages to perform geospatial analysis: 'RSAGA' [@brenningRSAGASAGAGeoprocessing2018], 'sf' [@pebesmaSimpleFeaturesStandardized2018] 'raster' [@hijmansRasterGeographicData2020] 'stars' [@pebesmaStarsSpatiotemporalArrays2020], and 'exactextractr' [@bastonExactextractrFastExtraction2020].

### Statistical analysis

We developed robust regression models with an M-estimation method to examine the effects of macro-scale drivers on diversity metrics. The robust regression analysis is insensitive to outliers of a response variable; therefore, this method allows us to account for uncertainties of our diversity estimates (see **Fish community data**). Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $w$, $log_{10}~y_w$, were assumed to follow a normal distribution as $log_{10}~y_w \sim Normal(\mu_w, \sigma^2)$. The parameter $\mu_w$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~WA$), log-transformed branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, the fraction of agricultural land use (logit-transformed), and dam density. Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land use variables to control their effects statistically. We did not include the fractions of urban and forest land use in the statistical models because these variables were strongly correlated with the fraction of agricultural land use (**Figure S19**). Similarly, we did not use mean elevation in the statistical models because it showed a moderate correlation with air temperature (**Figure S19**). Correlations for other combinations of explanatory variables were weak to moderate.
